---
title: "crash report (data clean)"
author: "Ava Haghighi"
format: html
editor: visual
---

## What do Montgomery County drivers get distracted by most often?

Based on the Montgomery crash data, a variable that categorizes the causes of driver distractions into more than 20groups has been broken down and grouped into key categories. These categories include: cellphone distractions, physical activity distractions, distractions caused by others in the vehicle, distractions due to lost in thought, and other distractions.

```{r importing data }
setwd("C:/Users/ava/Downloads/Data 205/final project")
```

```{r}
data <- read.csv("CrashReporting2.csv")
```

```{r loading librarys}
library(ggplot2)
library(dplyr)
library(tidyr)
library(summarytools)
```

```{r cleaning data}
data <- data %>%
  mutate(across(where(is.character), 
                ~ na_if(na_if(na_if(na_if(na_if(., ""), "N/A"), "NULL"), "Unknown"), "UNKNOWN"))) %>%


  mutate(across(where(is.character), 
                ~ na_if(., ""))) %>%

  mutate(across(where(is.numeric), 
                ~ na_if(., NA))) %>%
  mutate(across(where(is.integer), 
                ~ na_if(., NA)))

```


```{r}
# indicating how many groups of distraction there is in the data 
  unique_groups <- unique(data$Driver.Distracted.By)

print(unique_groups)
```


```{r Distraction grouped }
# dividing 23 groups in 5 main categorize 
distractions_grouped <- data.frame(distraction = data$Driver.Distracted.By, stringsAsFactors = FALSE)

distractions_grouped <- distractions_grouped %>%
  mutate(group = case_when(
    # Group 1: Not Distracted (for anything related to not being distracted)
    distraction %in% c("NOT DISTRACTED", "Not Distracted") ~ "Not Distracted",
    
    # Group 2: Distractions from Other Occupants (for distractions caused by other people in the vehicle)
    distraction %in% c("BY OTHER OCCUPANTS", "DISTRACTED BY OUTSIDE PERSON OBJECT OR EVENT", "BY MOVING OBJECT IN VEHICLE") ~ "Distractions from Other Occupants",
    
    # Group 3: Using Devices (for distractions related to using devices like phones, navigation systems, etc.)
    distraction %in% c("USING DEVICE OBJECT BROUGHT INTO VEHICLE", 
                       "OTHER ELECTRONIC DEVICE (NAVIGATIONAL PALM PILOT)", 
                       "USING OTHER DEVICE CONTROLS INTEGRAL TO VEHICLE", 
                       "OTHER CELLULAR PHONE RELATED", 
                       "TALKING OR LISTENING TO CELLULAR PHONE", 
                       "DIALING CELLULAR PHONE", 
                       "TEXTING FROM A CELLULAR PHONE", 
                       "Manually Operating (dialing, playing game, etc.)") ~ "Using Devices",
    
    # Group 4: Inattention or Lost in Thought (for distractions due to inattention or thinking)
    distraction %in% c("INATTENTIVE OR LOST IN THOUGHT", "LOOKED BUT DID NOT SEE", 
                       "Other Action (looking away from task, etc.)", "Talking/listening") ~ "Inattention or Lost in Thought",
    
    # Group 5: Eating, Smoking, or Other Physical Distractions (for eating, smoking, or other actions)
    distraction %in% c("EATING OR DRINKING", "SMOKING RELATED") ~ "Eating, Smoking, or Other Physical Distractions",
    
    # Fallback for any other cases (if any value does not fit into above categories)
    TRUE ~ "Other"
  ))
```

```{r}
unique(distractions_grouped$group)
```


```{r}
distractions_grouped_filtered <- distractions_grouped %>%
  filter(!group %in% c("Not Distracted", "Other"))
```

```{r}
ggplot(distractions_grouped_filtered, aes(x = group)) + 
  geom_bar(fill = "red", color = "black") +
  labs(title = "Distribution of Driver Distraction Groups",
       x = "Distraction Group",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

